Exploring the Utility of Machine Learning Across Varied Data Formats

### 2024-05-25 homework2/paper/paper.tex
finished with logistic regression

\begin{document}
%
\title{Exploring the Utility of Machine Learning Across Varied Data Formats}
%
\author{Cărămidă Iustina-Andreea - 332CA}
%
\institute{Faculty of Automatic Control and Computer Science \\
University Politehnica of Bucharest \\
\email{iustina.caramida@stud.acs.upb.ro}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    This study investigates the applicability of machine learning techniques on 
    diverse datasets. We explore the effectiveness of two algorithms, Linear 
    Regression and Multi-Layered Perceptron (MLP), on predicting both health 
    outcomes and financial well-being. Specifically, we utilize a stroke 
    prediction dataset to assess the model's ability to identify individuals at 
    risk of stroke. Additionally, we employ a salary prediction dataset to 
    evaluate the model's capacity to classify individuals earning above a 
    specific income threshold (e.g., \$50,000 per year). Through comparative 
    analysis, this research aims to elucidate the strengths and limitations of 
    each algorithm when applied to these contrasting data types, offering 
    insights into their suitability for various prediction tasks. Furthermore, 
    we present a framework for data analysis, outlining essential steps for data
    cleaning, exploration, and preparation, which can be applied to enhance the 
    effectiveness of machine learning models across diverse datasets.

\keywords{\textit{Machine Learning \and Heterogeneous Data \and Comparative Analysis \and
Prediction Modeling \and Data Analysis Techniques \and Stroke Prediction \and
Salary Prediction \and Linear Regression \and Multi-Layered Perceptron \and 
DataPreprocessing} }

@@

\subsection{Methodology: Unveiling the Algorithms - Linear Regression and Multi-Layered Perceptron}
To investigate the performance on these contrasting datasets, this study employs 
two prominent ML algorithms: Linear Regression and Multi-Layered Perceptron (MLP).

Linear Regression is a well\-established technique known for its interpretability
and efficiency in uncovering linear relationships between features (data points) 
and target variables (what we want to predict). This makes it a valuable tool 
for understanding the underlying factors influencing a particular outcome, such 
as the relationship between blood pressure and stroke risk. However, its strength 
lies in capturing linear relationships. If the underlying relationships in the 
data are more complex and non-linear, then Linear Regression might not be as 
effective.

On the other hand, Multi-Layered Perceptrons (MLPs) are a type of artificial 
neural network capable of learning complex, non-linear patterns within data. 
Unlike Linear Regression, MLPs are not limited by linearity and can potentially 
capture more intricate relationships between features and target variables. This 
capability makes them particularly suitable for datasets with complex underlying 
structures, such as the factors influencing an individual's salary, which might 
involve a combination of education, experience, industry, and other factors 
interacting in non-linear ways.

\subsection{Research Objectives: Evaluating Algorithms, Unveiling Strengths and Weaknesses}
By comparatively analyzing the performance of Linear Regression and MLP on 
stroke and salary prediction tasks, this research seeks to achieve several key 
objectives:

\subsubsection{Evaluate the Suitability of Algorithms for Diverse Data Types:}
This involves assessing the effectiveness of each algorithm in capturing the 
underlying relationships within the stroke prediction and salary prediction 
datasets. We will determine which algorithm performs better on each dataset, 
offering insights into their suitability for different data types, such as 
linear datasets (blood pressure and stroke risk) versus potentially non-linear 
datasets (factors influencing salary).

\subsubsection{Gain Insights into Algorithmic Strengths and Weaknesses:}
By analyzing the comparative performance, we aim to highlight the scenarios where
each algorithm excels and identify areas where one might outperform the other. 
This will provide valuable guidance for researchers and practitioners in 
selecting the most appropriate algorithm for their specific prediction tasks. 
For instance, if interpretability is crucial (e.g., understanding the factors 
influencing stroke risk), Linear Regression might be preferred. If the data is 
likely to have complex, non-linear relationships, then an MLP might be a better 
choice.

@@

% ------------------------------------------------------------------------------
\section{Algorithms Designs}
Algorithm design is a critical aspect of computer science and machine learning, 
focusing on creating efficient and effective methods to solve complex problems. 
The process involves the careful selection of algorithms based on the specific 
characteristics of the data and the desired outcomes. This document explores the 
application of two prominent machine learning algorithms, Linear Regression and 
Multi-Layered Perceptron (MLP), on diverse datasets. The goal is to compare their 
performance and suitability for different types of prediction tasks, particularly 
in the contexts of stroke prediction and salary prediction.

\subsection{Linear Regression}
Linear Regression is one of the most fundamental and widely used algorithms in 
machine learning. It is known for its simplicity, interpretability, and 
efficiency in identifying linear relationships between variables. The primary 
purpose of Linear Regression is to model the relationship between a dependent 
variable and one or more independent variables by fitting a linear equation to 
observed data.

In the context of stroke prediction, Linear Regression can be used to analyze 
how various factors such as age, blood pressure, cholesterol levels, and smoking 
history influence the likelihood of experiencing a stroke. Its straightforward 
approach allows for clear insights into the significance and impact of each 
factor on stroke risk.

In the \textit{linear\_regression} folder at the root of the project directory,
we implemented the Linear Regression algorithm in two different ways:

\begin{itemize}
    \item \textit{Linear Regression with Scikit-Learn:} We used the Scikit-Learn
    library to implement Linear Regression on the preprocessed datasets.
    \item \textit{Linear Regression from Scratch:} We implemented Linear Regression
    as in the 8th laboratory from the Artificial Intelligence course.
\end{itemize}

I choose to implement the RidgeRegression algorithm. Because of the large number of
features in the dataset, the RidgeRegression algorithm is more suitable than the
LinearRegression algorithm. The RidgeRegression algorithm is a regularized version
of Linear Regression that adds a penalty term to the loss function to prevent overfitting.
This penalty term helps to reduce the model's complexity and improve its generalization
performance. The alpha parameter controls the strength of the regularization, with
higher values of alpha leading to stronger regularization. I used the value of alpha
1.75 because it provided the best results in the experiments.

@@

The results of both implementations for each dataset are saved in the \textit{LinearRegression}
folder at the specific dataset's root.

>>>>

\begin{document}
%
\title{Exploring the Utility of Machine Learning Across Varied Data Formats}
%
\author{Cărămidă Iustina-Andreea - 332CA}
%
\institute{Faculty of Automatic Control and Computer Science \\
University Politehnica of Bucharest \\
\email{iustina.caramida@stud.acs.upb.ro}
}
%
\maketitle              % typeset the header of the contribution
%
\begin{abstract}
    This study investigates the applicability of machine learning techniques on 
    diverse datasets. We explore the effectiveness of two algorithms, Logistic 
    Regression and Multi-Layered Perceptron (MLP), on predicting both health 
    outcomes and financial well-being. Specifically, we utilize a stroke 
    prediction dataset to assess the model's ability to identify individuals at 
    risk of stroke. Additionally, we employ a salary prediction dataset to 
    evaluate the model's capacity to classify individuals earning above a 
    specific income threshold (e.g., \$50,000 per year). Through comparative 
    analysis, this research aims to elucidate the strengths and limitations of 
    each algorithm when applied to these contrasting data types, offering 
    insights into their suitability for various prediction tasks. Furthermore, 
    we present a framework for data analysis, outlining essential steps for data
    cleaning, exploration, and preparation, which can be applied to enhance the 
    effectiveness of machine learning models across diverse datasets.

\keywords{\textit{Machine Learning \and Heterogeneous Data \and Comparative Analysis \and
Prediction Modeling \and Data Analysis Techniques \and Stroke Prediction \and
Salary Prediction \and Logistic Regression \and Multi-Layered Perceptron \and 
DataPreprocessing} }

@@

\subsection{Methodology: Unveiling the Algorithms - Logistic Regression and Multi-Layered Perceptron}
This section delves into the core methodologies employed in this study: 
Logistic Regression and Multi-Layered Perceptron (MLP). Both algorithms fall 
under the umbrella of supervised learning, where a model learns from labeled 
data to make predictions on unseen examples. Here, we unveil the underlying 
principles and functionalities of each technique.

Logistic regression serves as a foundational algorithm for classification tasks.
It establishes a mathematical model that maps input features (e.g., age, blood 
pressure) to a probability of a specific outcome (e.g., stroke occurrence). The 
model essentially learns a decision boundary, separating observations with high 
and low probabilities of the target outcome. This approach makes logistic 
regression well-suited for analyzing datasets like the stroke prediction one,
where the goal is to categorize individuals based on their risk level.

On the other hand, Multi-Layered Perceptron (MLP) represents a more complex 
architecture - a type of artificial neural network. It consists of interconnected 
layers of artificial neurons, mimicking the structure of the human brain. Each 
layer transforms the received data using activation functions, ultimately leading 
to an output prediction. MLP's strength lies in its ability to learn complex, 
non-linear relationships within data. This makes it a powerful tool for tackling 
intricate prediction problems, potentially outperforming logistic regression 
when the underlying relationships are not easily captured by a linear model. 
The salary prediction dataset, where various factors might influence income 
levels, could be a prime candidate for exploration with MLP.

\subsection{Research Objectives: Evaluating Algorithms, Unveiling Strengths and Weaknesses}
By comparatively analyzing the performance of Logistic Regression and MLP on 
stroke and salary prediction tasks, this research seeks to achieve several key 
objectives:

\subsubsection{Evaluate the Suitability of Algorithms for Diverse Data Types:}
This involves assessing the effectiveness of each algorithm in capturing the 
underlying relationships within the stroke prediction and salary prediction 
datasets. We will determine which algorithm performs better on each dataset, 
offering insights into their suitability for different data types.

\subsubsection{Gain Insights into Algorithmic Strengths and Weaknesses:}
By analyzing the comparative performance, we aim to highlight the scenarios where
each algorithm excels and identify areas where one might outperform the other. 
This will provide valuable guidance for researchers and practitioners in 
selecting the most appropriate algorithm for their specific prediction tasks.

@@

% ------------------------------------------------------------------------------
\section{Algorithms Designs}
Algorithm design is a critical aspect of computer science and machine learning, 
focusing on creating efficient and effective methods to solve complex problems. 
The process involves the careful selection of algorithms based on the specific 
characteristics of the data and the desired outcomes. This document explores the 
application of two prominent machine learning algorithms, Logistic Regression and 
Multi-Layered Perceptron (MLP), on diverse datasets. The goal is to compare their 
performance and suitability for different types of prediction tasks, particularly 
in the contexts of stroke prediction and salary prediction.

\subsection{Logistic Regression}
Logistic regression is a fundamental statistical method employed for classification 
tasks in machine learning. It establishes a mathematical model that maps a set
of input features (independent variables) to a probability of a specific outcome 
(dependent variable). The core functionality lies in estimating the odds of a 
particular class membership (e.g., presence of stroke) based on the input 
features. The resulting model essentially learns a decision boundary, 
separating observations with high and low probabilities of belonging to the 
target class. This characteristic makes logistic regression particularly 
well-suited for analyzing datasets where the outcome variable is binary 
(e.g., stroke occurrence vs. no stroke occurrence).

In the \textit{logistic\_regression} folder at the root of the project directory,
we implemented the Logistic Regression algorithm in two different ways:

\begin{itemize}
    \item \textit{Logistic Regression with Scikit-Learn:} We used the Scikit-Learn
    library to implement Logistic Regression on the preprocessed datasets.
    \item \textit{Logistic Regression from Scratch:} We implemented Logistic Regression
    from scratch using the Negative Log-Likelihood method and the Gradient Descent
    optimization algorithm.
\end{itemize}

Before starting the implementation of the Logistic Regression algorithm, we need to
encode the categorical attributes in the datasets. Categorical attributes are non-numeric
attributes that represent discrete categories or groups. These attributes need to be
encoded into a numerical format before they can be used in machine learning algorithms.

@@

In the Logistic Regression with Scikit-Learn implementation, we used the \textit{LogisticRegression}
class from the \textit{sklearn.linear\_model} module to train the model on the preprocessed
datasets, without setting any hyperparameters, using the default values.
For the Logistic Regression from Scratch implementation, we implemented the Negative Log-Likelihood
loss function and the Gradient Descent optimization algorithm. We trained the model on the
preprocessed datasets, setting the learning rate to 0.01 and the number of epochs to 10000. For the
regularization, we used the Ridge Regression technique.

The results of both implementations for each dataset are saved in the \textit{LogisticRegression}
folder at the specific dataset's root.
